Towards typed repositories of proofs

### 2010-07-05 talks/MIPS2010/slides.tex
bilan de la journée: niet!
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@136 403e655e-980a-47bd-8c60-1ee5348bb6b7

\documentclass[red]{beamer}
\usepackage{ucs}
\usepackage{mathpartir}
\usepackage{amsfonts,amsmath,amscd}
\usepackage{stmaryrd}
\usepackage[utf8]{inputenc}
\usepackage[protrusion=true,expansion=true]{microtype}

@@

\setbeamertemplate{footline}
{ \sf
  \hbox{%
  \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.25ex,dp=1ex,left]{button}%
    \usebeamerfont{title in head/foot} MIPS 2010
  \end{beamercolorbox}%
  \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.25ex,dp=1ex,right]{frame
    number}%
    \usebeamerfont{date in head/foot}
    \insertframenumber{} / \inserttotalframenumber\hspace*{2ex} 
  \end{beamercolorbox}}%
  \vskip0pt%
}

%\setbeamertemplate{footline}[frame number]
\usenavigationsymbolstemplate{}

@@

\frame{\titlepage}

\section*{Introduction}

\frame{un \pause deux}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}
\section{Section 1}
\frame{bla}
\section{Section 2}
\frame{blo}

>>>>

\documentclass[ignorenonframetext,red]{beamer}

% \documentclass{article} \usepackage{beamerarticle}

\usepackage{ucs}
\usepackage{mathpartir}
\usepackage{amsfonts,amsmath,amscd}
\usepackage{stmaryrd}
\usepackage[utf8x]{inputenc}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{dot2texi}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}

@@

\setbeamertemplate{footline}[frame number]
\setbeamertemplate{navigation symbols}{}

@@

\begin{frame}
  \titlepage
  \mode<article>{
    \newcommand\url\texttt
    \maketitle}
\end{frame}
\section*{Introduction}

This talk will be about some remarks and directions on the way the
edition of formal proof is done in proof assistants.

\begin{frame}{A paradoxical situation}
  Most time is spent \emph{editing}, not \emph{writing} proofs.

  Workflow of formal mathematics is largely inspired by software
  development:
  \begin{itemize}
  \item Text-based
  \item Separate compilation (\textsf{make})
  \item Versioning (\textsf{cvs}, \textsf{svn}, \textsf{diff}s\ldots)
  \end{itemize}
\end{frame}

Let me start with two simple observations relating the activity of
both the working mathematician and the programmer. A first, obvious
fact is that both spend more time \emph{editing} than \emph{writing}:
their workflow is highly non-linear, involving experiments, fixes,
backtrack on previous modifications\ldots, eventually validated or
invalidated by some criteria like the absence of bugs or the validity
of a proof.

The second observation is that since the advent of modern proof
assistants, large bodies of mathematics have been formalized and the
way people managed these developments is very similar to the one used
in software development: we write proofs in text buffers, which are
linear, and launch proof-checking when we are done. To avoid
re-checking the whole development, we split it into mutually dependent
files and re-check only those who have changed. To keep track of old
versions, we use line-based version control systems that have no
awareness of the actual content of the files. The separation into
multiple files then serves two different purpose, and the fact that
they coincide is pure contingency: the purpose of splitting a
development into logical units, and the purpose of easing the
non-linear process of proof editing. Also, recording the actual text
of the proofs for versioning is often non-satisfactory: on one side it
contains mostly non-relevant informations (line breaks, spaces\ldots),
on the other the relevant ones, the validity of the proof, the
dependency between objects, is absent.

The paradox in this situation is that these very tools can be used to
reason about formal languages and especially proof languages. Couldn't
we adapt the methods developped to reason about a formal language to
the proof language of proof assistants itself? Could we replace this
legacy toolchain and make it language-aware?

\begin{frame}{Expectations}

\end{frame}

\begin{frame}{\mode<beamer>{Outline}}
  \tableofcontents
\end{frame}

\section{}

\subsection{Practical applications}

\section{A core meta-language for proof repositories}

\begin{frame}{Inspiration: the \textsf{git} storage model}
  \begin{itemize}
  \item ``Content-adressable''
  \item Name reflects content
  \item Maximal sharing (or hashconsing)
  \end{itemize}
\end{frame}

\begin{frame}{Incremental type-checking}
  A small language for representing proofs:
  \only<1>{\[ t\ ::=\ (x:t)\cdot t\ |\ x\ |\ t\ t\ |\ *\] \[\]}%
  \only<2>{\[ t\ ::=\ (x:t)\cdot t\ |\ \alert x\ |\ \alert{t\ t}\ |\ *\] \[\]}%
  \only<3>{\[ t\ ::=\ (x:t)\cdot t\ |\ \alert a\ |\ *\]%
    \[ a\ ::=\ x\ |\ a\ x \]} 
  \only<4>{\[ t\ ::=\ (x:t)\cdot t\ |\ a\
    |\ *\ |\ \alert{(x = a)\cdot t}\]%
    \[ a\ ::=\ x\ |\ a\ x \]}
  \begin{example}
    \[ \infer[Lam]{\Gamma,A\vdash B}{\Gamma\vdash A\to B} \]
    \[ \textsc{Lam}\ :\ (\Gamma : env)\cdot(A B : type)\cdot
    (d : \Gamma,A\vdash B)\cdot \Gamma\vdash A\to B \]
  \end{example}
\end{frame}
